\pkg{casebase}: An Alternative Framework For Survival Analysis and Comparison of Event Rates

### 2020-09-21 analysis/paper/paper.Rmd
cleaned CS3

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  cache = TRUE, 
  comment = "#>",
  fig.path = "../figures/",
  out.width = "\\textwidth",
   out.extra = "keepaspectratio=true"
)

@@

With the goal of fitting smooth-in-time hazard functions, Hanley & Miettinen [-@hanley2009fitting] proposed a general framework for estimating fully parametric hazard models via logistic regression. Their main idea is simple: comparing person-moments when the event of interest occurred with moments when patients were at risk. Their approach provides users familiar with generalized linear models a natural way of fitting parametric survival models. Moreover, their framework is very flexible: general functions of time can be estimated (e.g. using splines or general additive models), and hence these models retain some of the flexibility of Cox's partial likelihood approach. And since the unit of analysis is a person moment, time-varying covariates can also easily be included in this framework.

@@

To perform variable selection on the regression parameters $\theta \in \mathbb{R}^p$ of the hazard function, we can add a penalty to the likelihood and solve the following problem:
\begin{equation}
\min _{\theta \in \mathbb{R}^{p}}\,\,-\ell\left(\theta\right)+\sum_{j=1}^p w_j P(\theta_j;\lambda,\alpha)
\end{equation}
where $\ell\left(\theta\right)$ is the log of the likelihood function given in \eqref{eq:lik-function}, $P(\theta_j;\lambda,\alpha)$ is a penalty term controlled by the non-negative regularization parameters $\lambda$ and $\alpha$, and $w_j$ is the penalty factor for the $j$th covariate. These penalty factors serve as a way of allowing parameters to be penalized differently. For example, we could set the penalty factor for time to be 0 to ensure it is always included in the selected model.

@@

\begin{table}[ht]

@@

By splitting these tasks, we give flexibility to the user. While the method call in Step 2 allows further customization by using the \pkg{ggplot2} [@ggplot2] family of functions, users can use for the graphics system of their choice to create population-time plots from the object created in Step 1.

To illustrate these functions, we will use data from the European Randomized Study of Prostate Cancer Screening (ERSPC) [@schroder2009screening] which was extracted using the approach described in Liu *et al.* [-@liu2014recovering] This dataset is available through the \pkg{casebase} package. It contains the individual observations for 159,893 men  from seven European countries, who were between the ages of 55 and 69 years when recruited for the trial.

@@

  - \code{glm}: This is the familiar logistic regression.
  - \code{glmnet}: This option allows for variable selection using Lasso or elastic-net (cf. Case Study 3). This functionality is provided through the \pkg{glmnet} package [@friedman2010jss].
  - \code{gam}: This option provides support for *Generalized Additive Models* via the \pkg{mgcv} package [@hastie1987generalized].
  - \code{gbm}: This option provides support for *Gradient Boosted Trees* via the \pkg{gbm} package. This feature is still experimental.

@@

In the first two case studies, we described the basic functionalities of the \pkg{casebase} package: creating population-time plots, fitting parametric models for hazard functions, and estimating the corresponding cumulative incidence curves. For the third case study, we show how \pkg{casebase} can also be used for variable selection through regularized estimation of the hazard function. To do this, we simply replace logistic regression with a penalized counterpart. In \pkg{casebase}, we achieve this by using the \pkg{glmnet} package [@friedman2010jss].

To illustrate this functionality, we use the dataset from the Study to Understand Prognoses Preferences Outcomes and Risks of Treatment (SUPPORT) [@knaus1995support]. The SUPPORT dataset tracks death in five American hospitals within individuals who are considered seriously ill. The original data is available online from the Department of Biostatistics at Vanderbilt University [@harrell_2020]. The cleaned and imputed data consists of 9104 observations and 30 variables, and it is available as part of the \pkg{casebase} package. For more information about this dataset, the reader is encouraged to look at the documentation in our package.

In this section, we compare our penalized case-base approach to two other models: Cox regression, and penalized Cox regression (using the \pkg{glmnet} package). In all three models, we use all covariates except \code{sps} and \code{aps}. These two variables correspond to scores for predicting the outcome that were developed as part of the original study.

To compare the performance of our models, we first divide the data into a training and a test data, with 95% of the observations used for training all three models. To assess both discrimination and calibration, we use a time-dependent version of the classical Brier score that is adjusted for censoring [@graf1999ass]. The Brier score can be used with both parametric and semi-parametric models. We use the \pkg{riskRegression} to compute these scores for all models.

@@

For our penalized case-base model, we opt for the natural log of time; recall that this corresponds to a Weibull distribution. For fitting the penalized hazard, we use \code{fitSmoothHazard.fit}, which is a matrix interface to the function \code{fitSmoothHazard}. In order to use this interface, we first create a matrix \code{y} containing the time and event variables, and a matrix \code{x} containing all remaining covariates we would like to include in the model space. We used lasso penalization by setting \code{alpha = 1}. Moreover, we use the argument \code{penalty.factor} to give a regularization weight of 0 to the time variable, and therefore force it into the final model.

@@

Our first comparison plot examines the covariates that were selected between all three hazard functions. In \ref{fig:cs3lolliPlot}, we show the coefficient estimates for covariates that we selected by both penalized Cox and penalized case-base. From this plot, we can clearly see the shrinkage effect of using penalized estimation.

@@

Next, we compare the cumulative incidence estimation over the test set. The probabilities over time for each observation are averaged, resulting in the absolute risk curves in Figure \ref{fig:cs3FinalBrier} (A). We can see some minimal differences between the three models, with the Kaplan-Meier estimate giving a lower estimate than our three models. We also note that the apparent smoothness of the Cox and penalized Cox curves is due to the large number of observations in the training set, which is used to derive the Breslow estimate of the baseline hazard. As described above, we compare the performance between the models by computing the Brier scores over time. In Figure \ref{fig:cs3FinalBrier} (B), we can see that the Brier score is larger for the Kaplan-Meier estimate than for the other three models. On the other hand, the differences between these three models are minimal.

@@

In brief, case-base sampling can be used in conjunction with penalized logistic regression to perform variable selection with survival models.

@@

We would like to thank Yi Yang for helpful discussions on penalized regression models. \mbox{Bhatnagar} gratefully acknowledges funding via Discovery Grants from the Natural Sciences and Engineering Research Council of Canada (NSERC).

>>>>

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = FALSE,
  message = FALSE,
  echo = FALSE,
  cache = TRUE, 
  fig.pos = 'ht',
  comment = "#>",
  fig.path = "../figures/",
  out.width = "\\textwidth",
   out.extra = "keepaspectratio=true"
)

@@

With the goal of fitting smooth-in-time hazard functions, Hanley & Miettinen [-@hanley2009fitting] proposed a general framework for estimating fully parametric hazard models via logistic regression. Their main idea is simple: comparing person-moments when the event of interest occurred with moments when patients were at risk. Their approach handles censored data effortlessly and provides users familiar with generalized linear models a natural way of fitting parametric survival models. Moreover, their framework is very flexible: general functions of time can be estimated (e.g. using splines or general additive models), and hence these models retain some of the flexibility of Cox's partial likelihood approach. And since the unit of analysis is a person moment, time-varying covariates can also easily be included in this framework.

@@

To perform variable selection on the regression parameters $\theta \in \mathbb{R}^p$ of the hazard function, we can add a penalty to the likelihood and solve the following problem:
\begin{equation}
\min _{\theta \in \mathbb{R}^{p}}\,\,-\ell\left(\theta\right)+\sum_{j=1}^p w_j P(\theta_j;\lambda,\alpha) \label{eq:penest}
\end{equation}
where $\ell\left(\theta\right)$ is the log of the likelihood function given in \eqref{eq:lik-function}, $P(\theta_j;\lambda,\alpha)$ is a penalty term controlled by the non-negative regularization parameters $\lambda$ and $\alpha$, and $w_j$ is the penalty factor for the $j$th covariate. These penalty factors serve as a way of allowing parameters to be penalized differently. For example, we could set the penalty factor for time to be 0 to ensure it is always included in the selected model.

@@

\begin{table}[H]

@@

By splitting these tasks, we give flexibility to the user. While the method call in Step 2 allows further customization by using the \pkg{ggplot2} [@ggplot2] family of functions, users may choose the graphics system of their choice to create population-time plots from the object created in Step 1.

To illustrate these functions, we will use data from the European Randomized Study of Prostate Cancer Screening (ERSPC) [@schroder2009screening] which was extracted using the approach described in Liu *et al.* [-@liu2014recovering]. This dataset is available through the \pkg{casebase} package. It contains the individual observations for 159,893 men  from seven European countries, who were between the ages of 55 and 69 years when recruited for the trial.

@@

  - \code{glm}: This is the familiar logistic regression.
  - \code{glmnet}: This option allows for variable selection using the elastic-net [@zou2005regularization] penalty (cf. Case Study 3). This functionality is provided through the \pkg{glmnet} package [@friedman2010jss].
  - \code{gam}: This option provides support for *Generalized Additive Models* via the \pkg{mgcv} package [@hastie1987generalized].
  - \code{gbm}: This option provides support for *Gradient Boosted Trees* via the \pkg{gbm} package. This feature is still experimental.

@@

In the first two case studies, we described the basic functionalities of the \pkg{casebase} package: creating population-time plots, fitting parametric models for hazard functions, and estimating the corresponding cumulative incidence curves. For the third case study, we show how \pkg{casebase} can also be used for variable selection through regularized estimation of the hazard function as given by \eqref{eq:penest}.

To illustrate this functionality, we use the dataset from the Study to Understand Prognoses Preferences Outcomes and Risks of Treatment (SUPPORT) [@knaus1995support]. The SUPPORT dataset tracks death in five American hospitals within individuals who are considered seriously ill. The original data is available online from the Department of Biostatistics at Vanderbilt University [@harrell_2020]. The cleaned and imputed data consists of 9104 observations and 30 variables, and it is available as part of the \pkg{casebase} package. In the comparisons below, all covariates except \code{sps} and \code{aps} were modeled. These two variables correspond to scores for predicting the outcome that were developed as part of the original study. For more information about this dataset, the reader is encouraged to look at the documentation in our package.

For our penalized case-base model, we opted for the natural log of time which corresponds to a Weibull distribution. For fitting the penalized hazard, we use \code{fitSmoothHazard.fit}, which is a matrix interface to the \code{fitSmoothHazard} function. We must supply both a matrix \code{y} containing the time and event variables, and a matrix \code{x} containing all other covariates. We applied the lasso penalty by setting \code{alpha = 1} and assigned a \code{penalty.factor} ($w_j$) of 0 to the time variable to ensure it was in the selected model. We compared our approach to both Cox regression, and lasso penalized Cox regression (fitted via the \pkg{glmnet} package).

To compare the performance of our models, we split the data into 95% training and 5% test sets. To assess both discrimination and calibration, we used a time-dependent version of the classical Brier score that is adjusted for censoring [@graf1999ass]. The Brier score can be used with both parametric and semi-parametric models. We used the \pkg{riskRegression} package to compute these scores for all models.

@@

@@

In Figure \ref{fig:cs3lolliPlot}, we show the coefficient estimates for covariates that we selected by both penalized Cox and penalized case-base. We note that both penalized approaches produce similar results, and can clearly see the shrinkage effect owing to the $\ell_1$ penalty.

@@

We then compared the cumulative incidence estimation over the test set. The probabilities over time for each observation are averaged, resulting in the absolute risk curves shown in Figure \ref{fig:cs3FinalBrier} (A). We can see some minimal differences between the three models, with the Kaplan-Meier giving the lowest estimates across follow-up-time. Note that the apparent smoothness of the Cox and penalized Cox curves is due to the large number of observations in the training set, which is used to derive the Breslow estimate of the baseline hazard. As described above, we compare the performance between the models by computing the Brier scores over time. In Figure \ref{fig:cs3FinalBrier} (B), we can see that the Brier score is larger for the Kaplan-Meier estimate than for the other three models. On the other hand, the differences between these three models are minimal.

@@

In this third case study, we showed how case-base sampling can be used in conjunction with penalized logistic regression to perform variable selection in survival models.

@@

We would like to thank Yi Yang for helpful discussions on penalized regression models. \mbox{Bhatnagar} gratefully acknowledges funding via a Discovery Grant from the Natural Sciences and Engineering Research Council of Canada (NSERC).
